A Plasma Torus Around a Young Low-Mass Star

### 2025-04-04 paper/ms.tex
update

Given that a full radiative transfer simulation was beyond our scope,
we opted to construct a phenomenological model aimed at capturing 
the emission from the circumstellar material.  We did this by
fitting each spectral epoch with a multi-component
gaussian, after having subtracted the time-average line profile as in
Figure~\ref{fig:spec}e.  The results of this exercise are shown in
Figure~\ref{fig:halphamodel}; details in its implementation and
interpretation follow.

To the implement the model, we assumed that the ``inner'' ($K_{\rm
inner}$$\approx$2.5\,$v_{\rm eq}$) clump would be well-fit by a sum of
two gaussians because it visually appears double-peaked in the raw
data from $\phi$=0.15-0.35 and $\phi$=0.65-0.85
(Figure~\ref{fig:spec}b).  We assumed the ``outer'' ($K_{\rm
outer}$$\approx$3.9\,$v_{\rm eq}$) clump would be better fit by a
single gaussian, based on its behavior from $\phi$=0.6-0.9.  Each
gaussian component has three free parameters at each spectral epoch:
the mean $\mu$, standard deviation $\sigma$, and amplitude $A$.
Denote the two gaussians for the inner component $i=\{ 0, 1 \}$,
and the single outer component with $i=2$.  Given the complexity of the
line profile data (Figure~\ref{fig:halphamodel} left column), the
likelihood function for this model is multimodal.  We therefore
imposed the prior constraints that $A_i \sim \mathcal{U}[0,1]$,
$\sigma_i/v_{\rm eq} \sim \mathcal{U}[0,1]$, and further assumed
$K_i(t) \sim
\mathcal{U}[
  K_{\rm inner}\sin(\phi(t)) - v_{\rm eq},
  K_{\rm inner}\sin(\phi(t)) + v_{\rm eq}
]$
for the inner two components, and
$K_i(t) \sim
\mathcal{U}[
  K_{\rm outer}\sin(\phi(t) + \pi) - 2v_{\rm eq},
  K_{\rm outer}\sin(\phi(t) + \pi) + 2v_{\rm eq}
]$
for the outer component.
This latter assumption constrains the mean velocity of each component
to be near the value suggested by visual inspection, but provides
sufficient flexibility for the model to deviate from the assumption of
a sinusoid if the data dictate such behavior.
We fitted each component to the data {\it independently} using scipy's
non-linear least squares \texttt{curve\_fit} method
\cite{Virtanen2020}, and scaled the resulting parameter covariance
matrix by a constant factor to match the sample variance of the
residuals.
The choice to fit the $i=\{ 0, 1 \}$, and $i=2$ components
independently generally did not affect the results compared to if we
have fit all three simultaneously, except for at $\phi=1.00$ (see
Figure~\ref{fig:halphamodel}).

The resulting mean velocities, widths, and amplitudes are given in
Table~\ref{tab:halphamodelparams}.
%FIXME TODO gotta get this table being better... and in particular
%gotta not quite signal where there is nothing.  (mask out transits +
%secondary eclipses....)

The main benefit of this model is that it is
sufficiently flexible to capture the observed variability in
circumstellar emission, and

quantify the exact
orbital periods, velocities, and velocity dispersions of these clumps.

Figure~\ref{fig:halphamodel}

functions

{\it Constraints on physical dimensions of emitting region}---

@@

\begin{table}
  \centering
  \begin{tabular}{lcr}
  \hline 
  \hline 
  Time [BJD$_{\rm TDB}$] & RV (\kms) & $\sigma_{\rm RV}$ (\kms) \\
  % from tables/tab_TIC141146667_rel_rv.tex and
  % tables/tab_TIC141146667_times_rounded.csv
  \hline 
  2460357.954919 & 2.73 & 5.86 \\
  2460357.965845 & -4.40 & 2.37 \\
  2460357.976770 & -0.19 & 2.64 \\
  2460357.987698 & 3.84 & 2.87 \\
  2460357.998619 & 7.53 & 7.53 \\
  2460358.009538 & -1.98 & 1.44 \\
  2460358.020462 & 1.02 & 1.21 \\
  2460358.031383 & 0.64 & 7.03 \\
  2460358.042306 & -2.91 & 2.71 \\
  2460358.053228 & 8.93 & 6.75 \\
  2460358.064154 & 5.95 & 8.84 \\
  2460358.075075 & -2.25 & 3.06 \\
  2460358.085996 & 1.84 & 1.34 \\
  2460358.096918 & 2.41 & 8.24 \\
  2460358.107839 & -7.04 & 3.94 \\
  2460358.118760 & -2.24 & 3.07 \\
  2460358.129683 & -2.83 & 7.55 \\
  2460358.140606 & -0.59 & 2.26 \\
  2460358.151527 & 1.84 & 2.91 \\
  2460358.162448 & 4.54 & 3.95 \\
  2460358.173368 & 6.21 & 12.14 \\
  \hline
  \end{tabular}
  \caption{TIC~141146667 radial velocities relative to the systemic velocity.}
  \label{tab:rv}
\end{table}

@@

  \caption{Multi-gaussian model fit results to H$\alpha$ line profiles.}

>>>>

Given that a full radiative transfer simulation was outside our scope,
we opted to construct a phenomenological model aimed at capturing 
the emission from the circumstellar material.  We did this by
fitting each spectral epoch with a multi-component
gaussian, after having subtracted the time-average line profile as in
Figure~\ref{fig:spec}e.  The results of this exercise are shown in
Figure~\ref{fig:halphamodel}; details in its implementation and
interpretation follow.

To the implement the model, we assumed that the ``inner'' ($K_{\rm
inner}$$\approx$2.5\,$v_{\rm eq}$) clump would be well-fit by a sum of
two gaussians because it visually appears double-peaked in the raw
data from $\phi$=0.15-0.35 and $\phi$=0.65-0.85
(Figure~\ref{fig:spec}b).  We assumed the ``outer'' ($K_{\rm
outer}$$\approx$3.9\,$v_{\rm eq}$) clump would be better fit by a
single gaussian, based on its behavior from $\phi$=0.6-0.9.  Each
gaussian component has three free parameters at each spectral epoch:
the mean $\mu$, standard deviation $\sigma$, and amplitude $A$.
Denote the two gaussians for the inner component $i=\{ 0, 1 \}$,
and the single outer component with $i=2$.  Given the complexity of the
line profile data (Figure~\ref{fig:halphamodel} left column), the
likelihood function for this model is multimodal.  We therefore
imposed the prior constraints that $A_i \sim \mathcal{U}[0,1]$,
$\sigma_i/v_{\rm eq} \sim \mathcal{U}[0,1]$, and further assumed
$\mu_i(t) \sim
\mathcal{U}[
  K_{\rm inner}\sin(\phi(t)) - v_{\rm eq},
  K_{\rm inner}\sin(\phi(t)) + v_{\rm eq}
]$
for the inner two components, and
$\mu_2(t) \sim
\mathcal{U}[
  K_{\rm outer}\sin(\phi(t) + \pi) - 2v_{\rm eq},
  K_{\rm outer}\sin(\phi(t) + \pi) + 2v_{\rm eq}
]$
for the outer component.  This prior on the means mitigates the
multimodality of the likelihood by requiring the mean velocity of each
component to be within a one or two $v_{\rm eq}$ of the time-variable
sinusoid suggested by visual inspection.  We fitted each component to
the data {\it independently} using scipy's non-linear least squares
\texttt{curve\_fit} implementation \cite{Virtanen2020}, and scaled the
resulting parameter covariance matrix by a constant factor to match
the sample variance of the residuals.  The resulting means,
amplitudes, and standard deviations for each component are given in
Extended Data Table~\ref{tab:halphamodelparams}.

Caution is required in interpreting this model's results.  At times
the model overfits, because sometimes the most significant spectral
feature around any component's prior is not statistically significant.
During such spectral epochs, e.g.~the ``outer'' clump at $\phi$=1.07,
the model fits noise, not signal.  At other times, the model
underfits.  For instance, the sudden blue rise near $\phi$=0.2 is not
well-described by a gaussian; the assumed functional form is one of
convenience.  Finally, the model fits the $i=\{ 0, 1 \}$, and $i=2$
components independently.  At $\phi=0.0$ and $1.0$,
Figure~\ref{fig:halphamodel} suggests that the emission
might come from either the inner or outer components.
Physically however, at this time the inner component
is in transit, and the outer component is in secondary eclipse and
would likely be blocked by the star.  At $\phi=0.0$, the double-peaked
emission profile also matches that seen shortly afterward (at
$\phi$=0.14) when the
inner clump is viewed off-disk.  A physical interpretation of the
model should therefore discard the outer clump results at this
particular epoch.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf7.pdf}
  \caption{Orbits fit to mean radial velocities (RVs) extracted from 
  H$\alpha$ profile fits in Figure~\ref{fig:halphamodel}.
  Each marker represents the measured best-fit mean at a given epoch.
  Solid circles were used in the fits; transparent X markers
  were excluded due to concerns about their epoch-by-epoch
  interpretation (see text).  The purpose of this exercise was to
  quantify the orbital periods and semi-amplitudes of the
  circumstellar clumps.  The inner double-peaked clump is shown in
  orange and blue; the outer clump is shown in green.  Five draws from
  each posterior probability distribution are plotted.}
  \label{fig:orbits}
\end{figure}

Taking these caveats into account, we used the results from Extended
Data Table~\ref{tab:halphamodelparams} to evaluate various models for
the orbits.  Our main goal in this exercise was to quantify the
orbital periods and velocity semi-amplitudes of the clumps.
Figure~\ref{fig:orbits} shows representative results for the
circular orbit case; we found no reason to extend the exercise to
eccentric orbits.
Based on visual inspection of the line profile fits, we manually excluded 
epochs where we do not believe our gaussian profile fitting provided a
reliable representation of the circumstellar emission.  We then performed
the orbit fits using NumPyro to sample over the Gaussian likelihood with the
NUTS algorithm \cite{Phan2019}.  When performing the fits, we adopted
the uncertainties on each estimated mean radial velocity value, and
added a jitter term in quadrature to inflate these nominal measurement
uncertainties.  This procedure yielded period and semi-amplitude
values for circular orbits of
$P_0$=$3.92\pm0.03$\,hr, $K_0/v_{\rm eq}$=2.07$\pm$0.04;
$P_1$=$3.92\pm0.06$\,hr, $K_1/v_{\rm eq}$=2.88$\pm$0.10;
and $P_2$=$3.35\pm0.35$\,hr, $K_1/v_{\rm eq}$=4.16$\pm$0.56.
The periods for the inner double-peaked clump are consistent
with the photometric 3.930$\pm$0.001\,hr period to a precision of
about two minutes.
The ``period'' for the outer clump is more ambiguous because the
H$\alpha$ data only support the idea of a periodic orbit of material
well-fit by gaussian emission from $\phi$$\approx$0.5-1.0.
From $\phi$=0-0.2, there is no detectable emission,
and from $\phi$=0.2-0.4, the suddenly-detectable emission spans
1-4\,$v_{\rm eq}$ without taking a clear gaussian shape.
While the outer-most edge of the $\phi$=0.2-0.4 emission provides a
closer match to the expectation of a circular orbit, 
the idea of invoking a particular functional form for this component 
seems fine-tuned.
We instead emphasize that although this emission is present, 
its variability in time is poorly explained by the idea of a
stable, long-lived clump of material.
Although additional observations would be needed to conclusively determine whether or not 
this component of the system
is long-lived, the lack of any photometric variability in the TESS
data during its expected transits might suggest that it is not.

{\it Constraints on physical dimensions of emitting region}---
%TODO FIXME TODO FIXME

@@

\begin{table}
  \centering
  \begin{tabular}{lcr}
  \hline 
  \hline 
  Time [BJD$_{\rm TDB}$] & RV (\kms) & $\sigma_{\rm RV}$ (\kms) \\
  % from tables/tab_TIC141146667_rel_rv.tex and
  % tables/tab_TIC141146667_times_rounded.csv
  \hline 
  2460357.954919 & 2.73 & 5.86 \\
  2460357.965845 & -4.40 & 2.37 \\
  2460357.976770 & -0.19 & 2.64 \\
  2460357.987698 & 3.84 & 2.87 \\
  2460357.998619 & 7.53 & 7.53 \\
  2460358.009538 & -1.98 & 1.44 \\
  2460358.020462 & 1.02 & 1.21 \\
  2460358.031383 & 0.64 & 7.03 \\
  2460358.042306 & -2.91 & 2.71 \\
  2460358.053228 & 8.93 & 6.75 \\
  2460358.064154 & 5.95 & 8.84 \\
  2460358.075075 & -2.25 & 3.06 \\
  2460358.085996 & 1.84 & 1.34 \\
  2460358.096918 & 2.41 & 8.24 \\
  2460358.107839 & -7.04 & 3.94 \\
  2460358.118760 & -2.24 & 3.07 \\
  2460358.129683 & -2.83 & 7.55 \\
  2460358.140606 & -0.59 & 2.26 \\
  2460358.151527 & 1.84 & 2.91 \\
  2460358.162448 & 4.54 & 3.95 \\
  2460358.173368 & 6.21 & 12.14 \\
  \hline
  \end{tabular}
  \caption{TIC~141146667 radial velocities relative to the systemic
  velocity based on the 7699\,\AA\ resonance line and TiO bandheads.}
  \label{tab:rv}
\end{table}

@@

  \caption{Best-fit parameters from multi-gaussian model fit to H$\alpha$ line
  profiles.  These results should be treated with caution; only a
  subset of the spectral epochs provided statistically significant
  detections of the circumstellar emission (see Extended Data
  Figure~\ref{fig:halphamodel}).}
